Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2023-10-09 paper/paper-infer.tex
rm 'new' for AJHG

%\linenumbers
%\doublespacing
\onehalfspacing
\usepackage[numbers,super]{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}

@@

LDpred2 is a widely used Bayesian method for building polygenic scores (PGS). LDpred2-auto can infer the two parameters from the LDpred model, the SNP heritability $h^2$ and polygenicity $p$, so that it does not require an additional validation dataset to choose best-performing parameters. 
The main aim of this paper is to properly validate the use of LDpred2-auto for inferring multiple genetic parameters.
Here, we present a new version of LDpred2-auto that adds an optional third parameter $\alpha$ to its model, for modeling negative selection. We then validate the inference of these three parameters (or two, when using the previous model). We also show that LDpred2-auto provides per-variant probabilities of being causal that are well calibrated, and can therefore be used for fine-mapping purposes. We also derive a new formula to infer the out-of-sample predictive performance $r^2$ of the resulting PGS directly from the Gibbs sampler of LDpred2-auto. Finally, we extend the set of HapMap3 variants recommended to use with LDpred2 with 37\% more variants to improve the coverage of this set, and show that this new set of variants captures 12\% more heritability and provides 6\% more predictive performance, on average, in UK Biobank analyses.

@@

Most traits and diseases in humans are heritable.
What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero effect on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
As for the distribution of causal effects relative to their allele frequencies, it is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
In this model, the expected phenotypic variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$. 
These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher,speed2020evaluating} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
Not all methods have the same modeling assumptions; for example, LDAK-SumHer assumes a different prior than SBayesS and LDpred2-auto, and does not estimate all the same parameters. It can estimate the SNP heritability and $\alpha$. However, it cannot estimate the polygenicity nor the per-variant probabilities of being causal (since it inherently assumes an infinitesimal model, i.e.\ $p = 1$). Moreover, since it does not sample effects, it cannot be used to estimate the predictive performance $r^2$ with the new formula we propose in this paper.

As previously shown by \citet{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $r^2_\text{max}$ and $r^2_{g\hat{g}}$ are the maximum achievable squared correlations between the genetic predictor and respectively the phenotype and its genetic component, $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants. 
Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
LDpred2 is a widely used polygenic score method that can directly build PGS using resulting summary statistics from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
LDpred2 is a Bayesian approach that uses the SNP heritability $h^2$ and polygenicity $p$ as parameters of its model. LDpred2-auto, one variant of LDpred2, can directly estimate these two parameters from GWAS summary statistics, making it applicable even when no validation data is available for tuning these two model parameters \cite[]{prive2020ldpred2}.

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.9\textwidth]{overview}}
	\caption{Overview of what LDpred2-auto can now provide. For individual CIs of polygenic scores, please refer to the work of \citet{ding2022large,ding2022polygenic}. HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend using for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$). CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting). \label{fig:overview}}
\end{figure}

@@

For simulations, we use the UK Biobank imputed (BGEN) data, read as allele dosages with function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{bycroft2018uk,prive2017efficient}. 
We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
Since we run lots of different models, we restrict the simulations to chromosomes 3, 6, 9, 12, 15, 18 and 21, resulting in a set of 322,805 SNPs.
We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020). These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in \citet{bycroft2018uk}).
To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.  
This results in 356,409 individuals of Northwestern European ancestry.
We randomly sample 200,000 individuals to form a training set (to run the GWAS), and use the remaining individuals to form a test set (to evaluate the predictive models).

@@

We construct and use the same phenotypes as in \citet{prive2021high}.
About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
The other half consists of phenotypes defined in UKBB fields based on manual curation \cite[]{prive2021high}.
As covariates, we first recompute PCs for the homogeneous subset of individuals previously defined using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}. We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates (to a total of eight).

We use the LD matrix with independent LD blocks computed in \citet{prive2021identifying}.
We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \citet{prive2021high} (``hm3\_altpop'').
We apply the optimal algorithm developed in \citet{prive2021optimal} to obtain independent LD blocks, as recommended in \citet{prive2021identifying}.
We finally define a fourth LD reference by extending the set of HapMap3 variants (see next Methods section) and using 20,000 individuals from the previously selected ones (``hm3\_plus'').

@@

The HapMap3 variants generally provide a good coverage of the whole genome.
We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ `A' and `T', or `C' and `G'), and which includes SNPs only (e.g.\ no indel).
Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
There are around 11.5M such variants.
Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
Finally, we design an algorithm which aims at maximizing the tagging of all these variants read. We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set, which we call HapMap3+.
We start by including all previously used HapMap3 variants.
Then, for the sake of simplicity, we use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
Note that we only allow non-ambiguous SNPs to be included.
This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \citet{prive2021optimal} to obtain 431 independent LD blocks.
Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (respectively 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).

@@

To infer the out-of-sample predictive performance $r^2$ (and its CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
Intuitively, if prediction is perfect then $\boldsymbol{\beta_1} = \boldsymbol{\beta_2}$ and $r^2 = h^2$; when power is very low, $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are almost uncorrelated and $r^2 \approx 0$.
Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
These previous works have shown that $r^2$ can be approximated by $\big(\boldsymbol{\bar{\beta}}^T \boldsymbol{\hat{\beta}}_\text{test}\big)^2 / \big(\boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}\big)$ where $\boldsymbol{\bar{\beta}}$ and $\boldsymbol{\hat{\beta}}_\text{test}$ are respectively the predictive effects from the training set and the marginal effects from the test set (after scaling).
Note that $E[\boldsymbol{\beta_1}] = \boldsymbol{\bar{\beta}}$, $E[\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}] = \boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}$, and, when the test sample has the same genetic ancestry as the GWAS used for training (to get the summary statistics), $E[\boldsymbol{R} \boldsymbol{\beta_2}] \approx \boldsymbol{\hat{\beta}}_\text{test}$.
Therefore, we propose to use $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ as a sample of $r^2$.
This can also be computed for a specific chain by taking two samples $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ that are far enough on the same chain to remove the possible autocorrelation. 
This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
Note that \citet{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly. Therefore picking two LDpred2-auto samples that are 100 iterations apart should be more than enough to ensure quasi-independence of these samples.

In this paper, we check this new $r^2$ approximation using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
These are compared to the partial-$r^2$ (on individual-level data from a separate test set). The partial correlation is computed with function \texttt{pcor} from R package bigstatsr, adjusting for the same eight covariates as in the GWAS, then squared (while keeping the sign). Corresponding 95\% confidence intervals are estimated through bootstrapping individual indices.

@@

For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP heritability $h^2$, the polygenicity $p$ (i.e.\ the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
This function first picks a proportion $p$ of causal variants at random, samples effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scales the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix. Finally, some Gaussian noise is added so that the final phenotype has a variance of 1. 
Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a random subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \citet{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the extended model including a third parameter $\alpha$ (using option \texttt{use\_MLE}, Methods). 
LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale). Then some of these chains are filtered out for quality control (Methods).

@@

We use the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations. 
To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using eight covariates (Methods).
We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the extended 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citet{prive2021identifying}).

@@

Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citet{prive2021identifying}.
Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure S31). $h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the N.W. European LD reference) and a standard sampling can be used in this case (Figure S31).

@@

Here we run three LDpred2-auto models for height, one from the same 305K training UKBB individuals used before (with available height, out of 306K), one based on 100K UKBB individuals (as a random subset of the previous 305K), and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries \cite[]{yengo2022saturated}.
We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \citet{prive2021using}, and find that 81.9\% are from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi genetic ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
For this set of external GWAS summary statistics, we therefore use the same N.W. European LD matrix as used in UKBB analyses. 
Note that we use the HapMap3+ set of 1,444,196 SNPs here, however, for the GWAS meta-analysis, only 1,013,499 SNPs (out of 1,373,020) are overlapping with HapMap3+ and passing quality control.

@@

\vspace*{3em}
%\clearpage

>>>>

%\linenumbers
%\doublespacing
\onehalfspacing
\usepackage[numbers,super]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

@@

LDpred2 is a widely used Bayesian method for building polygenic scores (PGS). LDpred2-auto can infer the two parameters from the LDpred model, the SNP heritability $h^2$ and polygenicity $p$, so that it does not require an additional validation dataset to choose best-performing parameters. 
The main aim of this paper is to properly validate the use of LDpred2-auto for inferring multiple genetic parameters.
Here, we present a new version of LDpred2-auto that adds an optional third parameter $\alpha$ to its model, for modeling negative selection. We then validate the inference of these three parameters (or two, when using the previous model). We also show that LDpred2-auto provides per-variant probabilities of being causal that are well calibrated, and can therefore be used for fine-mapping purposes. We also introduce a formula to infer the out-of-sample predictive performance $r^2$ of the resulting PGS directly from the Gibbs sampler of LDpred2-auto. Finally, we extend the set of HapMap3 variants recommended to use with LDpred2 with 37\% more variants to improve the coverage of this set, and show that this new set of variants captures 12\% more heritability and provides 6\% more predictive performance, on average, in UK Biobank analyses.

@@

Most traits and diseases in humans are heritable.
What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero effect on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
As for the distribution of causal effects relative to their allele frequencies, it is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
In this model, the expected phenotypic variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$. 
These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher,speed2020evaluating} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
Not all methods have the same modeling assumptions; for example, LDAK-SumHer assumes a different prior than SBayesS and LDpred2-auto, and does not estimate all the same parameters. It can estimate the SNP heritability and $\alpha$. However, it cannot estimate the polygenicity nor the per-variant probabilities of being causal (since it inherently assumes an infinitesimal model, i.e.\ $p = 1$). Moreover, since it does not sample effects, it cannot be used to estimate the predictive performance $r^2$ with the formula we propose in this paper.

As previously shown by \citep{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $r^2_\text{max}$ and $r^2_{g\hat{g}}$ are the maximum achievable squared correlations between the genetic predictor and respectively the phenotype and its genetic component, $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants. 
Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
LDpred2 is a widely used polygenic score method that can directly build PGS using resulting summary statistics from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
LDpred2 is a Bayesian approach that uses the SNP heritability $h^2$ and polygenicity $p$ as parameters of its model. LDpred2-auto, one variant of LDpred2, can directly estimate these two parameters from GWAS summary statistics, making it applicable even when no validation data is available for tuning these two model parameters \cite[]{prive2020ldpred2}.

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.9\textwidth]{overview}}
	\caption{Overview of what LDpred2-auto can now provide. For individual CIs of polygenic scores, please refer to the work of \citep{ding2022large,ding2022polygenic}. HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend using for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$). CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting). \label{fig:overview}}
\end{figure}

@@

For simulations, we use the UK Biobank imputed (BGEN) data, read as allele dosages with function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{bycroft2018uk,prive2017efficient}. 
We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
Since we run lots of different models, we restrict the simulations to chromosomes 3, 6, 9, 12, 15, 18 and 21, resulting in a set of 322,805 SNPs.
We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020). These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in \citep{bycroft2018uk}).
To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.  
This results in 356,409 individuals of Northwestern European ancestry.
We randomly sample 200,000 individuals to form a training set (to run the GWAS), and use the remaining individuals to form a test set (to evaluate the predictive models).

@@

We construct and use the same phenotypes as in \citep{prive2021high}.
About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
The other half consists of phenotypes defined in UKBB fields based on manual curation \cite[]{prive2021high}.
As covariates, we first recompute PCs for the homogeneous subset of individuals previously defined using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}. We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates (to a total of eight).

We use the LD matrix with independent LD blocks computed in \citep{prive2021identifying}.
We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \citep{prive2021high} (``hm3\_altpop'').
We apply the optimal algorithm developed in \citep{prive2021optimal} to obtain independent LD blocks, as recommended in \citep{prive2021identifying}.
We finally define a fourth LD reference by extending the set of HapMap3 variants (see next Methods section) and using 20,000 individuals from the previously selected ones (``hm3\_plus'').

@@

The HapMap3 variants generally provide a good coverage of the whole genome.
We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ `A' and `T', or `C' and `G'), and which includes SNPs only (e.g.\ no indel).
Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
There are around 11.5M such variants.
Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
Finally, we design an algorithm which aims at maximizing the tagging of all these variants read. We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set, which we call HapMap3+.
We start by including all previously used HapMap3 variants.
Then, for the sake of simplicity, we use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
Note that we only allow non-ambiguous SNPs to be included.
This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \citep{prive2021optimal} to obtain 431 independent LD blocks.
Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (respectively 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).

@@

To infer the out-of-sample predictive performance $r^2$ (and its CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
Intuitively, if prediction is perfect then $\boldsymbol{\beta_1} = \boldsymbol{\beta_2}$ and $r^2 = h^2$; when power is very low, $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are almost uncorrelated and $r^2 \approx 0$.
Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
These previous works have shown that $r^2$ can be approximated by $\big(\boldsymbol{\bar{\beta}}^T \boldsymbol{\hat{\beta}}_\text{test}\big)^2 / \big(\boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}\big)$ where $\boldsymbol{\bar{\beta}}$ and $\boldsymbol{\hat{\beta}}_\text{test}$ are respectively the predictive effects from the training set and the marginal effects from the test set (after scaling).
Note that $E[\boldsymbol{\beta_1}] = \boldsymbol{\bar{\beta}}$, $E[\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}] = \boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}$, and, when the test sample has the same genetic ancestry as the GWAS used for training (to get the summary statistics), $E[\boldsymbol{R} \boldsymbol{\beta_2}] \approx \boldsymbol{\hat{\beta}}_\text{test}$.
Therefore, we propose to use $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ as a sample of $r^2$.
This can also be computed for a specific chain by taking two samples $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ that are far enough on the same chain to remove the possible autocorrelation. 
This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
Note that \citep{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly. Therefore picking two LDpred2-auto samples that are 100 iterations apart should be more than enough to ensure quasi-independence of these samples.

In this paper, we check this $r^2$ approximation using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
These are compared to the partial-$r^2$ (on individual-level data from a separate test set). The partial correlation is computed with function \texttt{pcor} from R package bigstatsr, adjusting for the same eight covariates as in the GWAS, then squared (while keeping the sign). Corresponding 95\% confidence intervals are estimated through bootstrapping individual indices.

@@

For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP heritability $h^2$, the polygenicity $p$ (i.e.\ the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
This function first picks a proportion $p$ of causal variants at random, samples effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scales the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix. Finally, some Gaussian noise is added so that the final phenotype has a variance of 1. 
Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a random subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \citep{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the extended model including a third parameter $\alpha$ (using option \texttt{use\_MLE}, Methods). 
LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale). Then some of these chains are filtered out for quality control (Methods).

@@

We use the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations. 
To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using eight covariates (Methods).
We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the extended 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citep{prive2021identifying}).

@@

Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citep{prive2021identifying}.
Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure S31). $h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the N.W. European LD reference) and a standard sampling can be used in this case (Figure S31).

@@

Here we run three LDpred2-auto models for height, one from the same 305K training UKBB individuals used before (with available height, out of 306K), one based on 100K UKBB individuals (as a random subset of the previous 305K), and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries \cite[]{yengo2022saturated}.
We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \citep{prive2021using}, and find that 81.9\% are from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi genetic ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
For this set of external GWAS summary statistics, we therefore use the same N.W. European LD matrix as used in UKBB analyses. 
Note that we use the HapMap3+ set of 1,444,196 SNPs here, however, for the GWAS meta-analysis, only 1,013,499 SNPs (out of 1,373,020) are overlapping with HapMap3+ and passing quality control.

@@

%\vspace*{3em}
\clearpage
